---
title: "Part B"
output: html_document
date: "2025-02-25"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(archive)
```


# Input data

```{r}
# Specify the file path
rar_file <- "Training.rar"

# Extract the contents
#unpacked_files <- archive::archive_extract(rar_file)
try(archive_extract("Training.rar"), silent = TRUE)
```


```{r}
# Función global para extraer el ID numérico del nombre de archivo
extract_id <- function(filename) {
  matches <- regmatches(basename(filename), regexpr("[0-9]+", basename(filename)))
  as.numeric(matches)
}

get_sorted_filenames <- function(){
  # Obtener todos los archivos .jpg
  image_files <- list.files("Training", pattern = "\\.jpg$", full.names = TRUE)
  sorted_indices <- order(sapply(image_files, extract_id))
  image_files[sorted_indices]
}

# Uso
image_files <- get_sorted_filenames()
image_ids <- sapply(image_files, extract_id)

# Lectura de imágenes y creación de la matriz M
library(OpenImageR)
image_rows  <- list()
for (file in image_files) {
  img <- readImage(file)  # Cargar imagen
  
  # Aplanar los canales RGB y concatenarlos en una sola fila
  red <- as.vector(img[,,1])
  green <- as.vector(img[,,2])
  blue <- as.vector(img[,,3])
  
  image_rows[[file]] <- c(red, green, blue)
}

M <- do.call(rbind, image_rows)
dim(M)

```
# Part A

As we compute the small var-cov matrix, the principal components are representing the variance among images. If we were computing the var-cov matrix in the regular way, we would be representing the variance in the pixels. Anyways we obtain the same nonzero eigenvalues.

```{r}
compute_pca <- function(data) {
  # Compute mean
  mean_m <- colMeans(data)
  
  # Subtract the mean from each row
  centered_matrix <- sweep(data, 2, mean_m, FUN = "-")

  n <- nrow(centered_matrix)
  var_cov_small <- (1 / (n - 1)) * (centered_matrix %*% t(centered_matrix))
  eigen_dec <- eigen(var_cov_small)
  
  V <- t(centered_matrix) %*% eigen_dec$vectors
  V <- apply(V, 2, function(v) v / sqrt(sum(v^2)))
  
  return(list(mean_obs = mean_m, P = eigen_dec$vectors, D = eigen_dec$values, V=V))
}

pca_results <- compute_pca(M)

mean_obs <- pca_results$mean_obs
P <- pca_results$P #eigenvectors in image space
D <- pca_results$D #eigenvalues
V <- pca_results$V #eigenvectors in pixel space normalized
```

# Variance explained by each component
```{r}
for (i in seq_along(D)) {
  cat("Principal component ", i, " explains ",D[i] / sum(D)*100, "% of the variance.\n")
}
```


Let's get the projection of the training images in the pca space

```{r}
#Y <- centered_matrix %*% V
Y <- sweep(M, 2, mean_obs, FUN = "-") %*% V

```

```{r}
reduced_Y <- Y[,1:10]
reduced_Y <- cbind(reduced_Y, class = image_ids)
```


```{r}
compute_fda <- function(Y) {
  # Y is an n x p matrix where each row is a flattened image.
  
  # Compute the overall mean of the data (each column is a pixel)
  m <- colMeans(Y)
  
  # number of people in the dataset and which ones
  people <- unique(reduced_Y[, "class"])
  k <- length(people)
  
  # Number of observations per person
  ni <- sapply(people, function(cl) sum(class_vector == cl))
  
  # Initialize matrices S_W and S_B
  S_W <- matrix(0, ncol(Y), ncol(Y))
  S_B <- matrix(0, ncol(Y), ncol(Y))
  
  for (classes in people) {
    # Filtrar datos de la clase actual
    Y_class <- Y[Y[, ncol(Y)] == 1, , drop = FALSE]
    mean_class <- colMeans(Y_class)
    mean_class <- mean_class[,-c('class')]
    # ACUERDATE DE BORRAR LA COLUMNA CLASE.
    
    for (meani in mean_class){
      S_B <- S_B + length(mean_class)*sum((meani - m)*t(meani-m))
    }
  }
}

```

```{r}
fda <- function(Y) {
  # Y is an n x p matrix obtained from a PCA where the last column has to contain the class.
  
  # We extract the class column and convert it to a factor
  class_vector <- Y[, ncol(Y)]
  class_vector <- as.factor(class_vector)
  
  # We extract the feature matrix (all columns except the last one)
  X <- Y[, -ncol(Y), drop = FALSE]
  
  # We compute the overall mean of the features
  m <- colMeans(X)
  
  # Determine the number of classes and the unique classes
  people <- unique(class_vector)
  
  # We obtain the number of features (dimensions) to select the size of S_W and S_B
  p <- ncol(X)
  
  # We initialize the within-class scatter matrix (S_W) and the between-class scatter matrix (S_B)
  S_W <- matrix(0, nrow = p, ncol = p)
  S_B <- matrix(0, nrow = p, ncol = p)
  
  # For each class, we compute the class mean and we accumulate contributions to S_B and S_W
  for (cl in people) {
    
    # Submatrix of observations for the current class 'cl'
    X_class <- X[class_vector == cl, , drop = FALSE]
    n_i <- nrow(X_class)
    
    # We compute the mean of the current class
    mean_class <- colMeans(X_class)
    
    # We update S_B for each class
    # We perform the outer product of (mean_class - m), weighted by n_i
    S_B <- S_B + n_i * ((mean_class - m) %*% t((mean_class - m)))
    
    # To Update S_W we need to access every observation j of the class one by one
    # We create a for loop which goes through each observation to perform the
    # sum of (x_j - mean_class)(x_j - mean_class)^T
    for (j in 1:n_i) {
      diff_j <- as.matrix(X_class[j, ] - mean_class)
      S_W <- S_W + diff_j %*% t(diff_j)
    }
  }
  
  # We compute the eigenvalues and eigenvectors for the matrix solve(S_W) %*% S_B
  eig_values <- eigen(solve(S_W) %*% S_B)
  
  # We obtain the sorted indices of the eigenvalues in decreasing order
  sort_eig <- order(eig_values$values, decreasing = TRUE)
  
  # We sort the eigenvectors and eigenvalues according these indices
  eigenvectors <- eig_values$vectors[, sort_eig]
  eigenvalues <- eig_values$values[sort_eig]
  
  return(list(overall_mean = m, P = eigenvectors, D = eigenvalues))
}
```


```{r}
fda_results <- compute_fda(reduced_Y)
```

```{r}
fda_results <- compute_fda(reduced_Y)
# The overall mean face (used for centering):
mean_face <- fda_results$overall_mean

# The matrix of Fisher discriminant directions (columns are the directions):
fisher_directions <- fda_results$P

# The eigenvalues (discriminative power for each direction):
discriminant_power <- fda_results$D
```


